Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-09-26 article.Rmd
Implement suggestions from writing group

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name from historical data in order to take into account how naming practices change over time and space. In contrast to existing methods of predicting gender, this method provides a higher level of accuracy and a precise statement of the likelihood that a name was a given gender. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we study gender and gatekeeping in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

@@

As humanities researchers grapple with larger and larger datasets they will increasingly be able to not just extract existing information from that data but to also infer additional information from it [Goldstone and Underwood 2014].^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. [Weingart 2014] Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance [Healy 2013].] Many of these datasets deal with lists of people: company payrolls, military rosters, passenger bookings, records of correspondence, or lists of published works. Some of these records might contain associated information such a person's address, age, or rank, while others contain little more than a list of personal names. Given large but sparse datasets, researchers have to use their ingenuity to derive additional kinds of information. What can researchers infer from just a name? On a basic level, first names often imply a person's gender. "Jane Fay" is almost certainly a woman, while her brother "John Fay" is almost certainly a man. Given a dataset matching names to genders, a straightforward computer program can easily infer their respective genders.

@@

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender [Butler 1990, Butler 1993]. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. This method does not presuppose that the gender binary imposed by governments and their wider societies reflect how an individual identifies oneself and is neutral as to what the categories "male" or "female" actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with it. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

@@

The method's real utility stems from being able to process larger datasets than a single name. Let's use, say, a list of editors from a college newspaper to illustrate how a researcher might apply it to their own data. The package's prediction function allows researchers to choose which reference datasets they would like to use and the range of years for making their predictions.^[See the package vignette for more details: `vignette("predicting-gender", package = "gender")`.]

@@

```{r, echo=TRUE, results='markup'}
Map(gender, editors$name, years = editors$birth_year) %>% 
  do.call(rbind.data.frame, .)
```

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period.

@@

The path from a dissertation to a book review in the *American Historical Association* is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the *AHR* editors for review. As of August 2014 the journal’s website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)."^[http://www.indiana.edu/~ahrweb/book_review_guide.html - retrieved August 25 2014]. Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face insidious gender discrimination.

@@

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of scarcity in their sources. Our method uses the abundant existing datasets to infer more information from scarce datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make predictions about gender and naming practices. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.

>>>>

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name from historical data in order to take into account how naming practices change over time and space. In contrast to existing methods of predicting gender, this method provides a higher level of accuracy and a precise statement of the likelihood that a name was a given gender, especially for datasets where names can be associated with dates. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we study gender and gatekeeping in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

@@

As humanities researchers grapple with larger and larger datasets they will increasingly be able to not just extract existing information from that data but to also infer additional information from it [Goldstone and Underwood 2014].^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. [Weingart 2014] Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance [Healy 2013].] Many of these datasets deal with lists of people: company payrolls, military rosters, passenger bookings, records of correspondence, or lists of published works. Some of these records might contain associated information such as a person's address, age, or rank, while others contain little more than a list of personal names. A common question in the history of American religion, for example, is whether more women than men participated in a church or religious group. A simple membership roll might contain names, but it is unlikely to identify the gender of the members. One might also ask a question about changing patterns in male and female authorship of books. Library catalogs contain enormous amounts of data about publication, but they do not record the sex of authors.

Given large but sparse datasets, researchers have to use their ingenuity to derive additional kinds of information. What can researchers infer from just a name? On a basic level, first names often imply a person's gender. "Jane Fay" is almost certainly a woman, while her brother "John Fay" is almost certainly a man. Given a dataset matching names to genders, a straightforward computer program can easily infer their respective genders.

@@

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender [Butler 1990, Butler 1993]. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies such as the U.S. Census Bureau. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. This method does not presuppose that the gender binary imposed by governments and their wider societies reflect how an individual identifies oneself and is neutral as to what the categories "male" or "female" actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with it. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

@@

The method's real utility stems from being able to process larger datasets than a single name. Let's use, say, a hypothetical list of editors from a college newspaper to illustrate how a researcher might apply it to their own data. The package's prediction function allows researchers to choose which reference datasets they would like to use and the range of years for making their predictions.^[See the package vignette for more details: `vignette("predicting-gender", package = "gender")`.]

@@

```{r, echo=TRUE, results='markup'}
Map(gender, editors$name, years = editors$birth_year) %>% 
  do.call(rbind.data.frame, .) %>% print.data.frame(row.names = FALSE)
```

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We have found that although women have acheived rough equality in terms of the number of dissertations written, there is still a significant gap between male and female authorship of books in history.

@@

The path from a dissertation to a book review in the *American Historical Association* is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the *AHR* editors for review. As of August 2014 the journal’s website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)."^[http://www.indiana.edu/~ahrweb/book_review_guide.html - retrieved August 25 2014]. Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face gender discrimination.

@@

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of scarcity in their sources. Our method uses the abundant existing datasets to infer more information from scarce datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make predictions about gender and naming practices. This method is potentially useful to anyone who has a dataset of names, especially if those names can be associated with particular dates. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.
